\subsection{\texorpdfstring{Ethics approval and consent to participate
}{Ethics approval and consent to participate }}\label{ethics-approval-and-consent}

The Institutional Review Boards (IRB) of Fox Chase Cancer Center approved the study.

\subsection{\texorpdfstring{Consent for publication
}{Consent for publication }}\label{consent-for-publication}

Consent for publication was obtained from all participants.

\subsection{\texorpdfstring{Availability of data and materials
}{Availability of data and materials }}\label{availability-of-data-and-materials}

See Data Availability section.

\subsection{\texorpdfstring{Competing Interests
}{Competing Interests }}\label{competing-interests}

The authors declare that they have no competing interests.

\subsection{\texorpdfstring{Funding }{Funding }}\label{funding}

This research was supported by new faculty funds from Fox Chase Cancer
Center and the Cancer Center Support Grant P30CA006927.

\subsection{\texorpdfstring{Authors Contributions
}{Authors Contributions }}\label{authors-contributions}

H.L. and N.V. conceptualized the study. P.F.E. designed the
transcriptome study. S.P. and J.M. performed transcriptome experiments.
G.Y. and C.Q. conducted histological analysis. T.K., K.C and H.L.
performed computational analyses and developed machine learning models.
H.L. implemented the visualization application. T.K., H.L., S.P., K.C.,
J.W., I.A., and N.V. contributed to manuscript writing. H.L. and N.V.
supervised the study. All authors read and approved the final
manuscript.

\subsection{\texorpdfstring{Acknowledgements
}{Acknowledgements }}\label{acknowledgements}

Not applicable.

\subsection{\texorpdfstring{Authors' information
}{Authors' information }}\label{authors-information}

Not applicable.
